### 2019-10-22 ThesisJo/2_background.tex
2019-10-22

% Related work can be found e.g. in ultrasound imaging, where Chen et. al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a}. Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix. 
% \\
\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
\caption{Method relations. This thesis explores three beamformers reconstructing images from acoustic echoes: time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated trial-and-error). Article IV presents background on synthesis of image templates from reconstructed images. }%
\label{2_fig_article_relation}
\end{figure}

@@

Any wave, regardless of its nature, is governed by a partial differential equation known as the \emph{wave equation}. For a linear, homogeneous (translation-invariant) and isotropic (rotation-invariant) medium its form is
%
\begin{align}\label{eq_wave_equation}
\nabla^2 f 
 = \frac{1}{c}\frac{\delta^2 f}{\delta t^2},
\end{align}
%\frac{\delta^2 f}{\delta x^2} + \frac{\delta^2 f}{\delta y^2} + \frac{\delta^2 f}{\delta z^2}
%
where $\nabla^2$ is the Laplace operator, $f$ is field strength and $c$ is wave propagation speed. In complex mediums $c$ may depend on signal strength (non-linear), position (non-homogeneous/refractive), angle (anisotropic), frequency (dispersive) or be non-real (lossy). Deep sea is acoustically non-homogeneous due to varying pressure, temperature and salinity, but has negligible dispersion and attenuation. Biological tissue both non-homogeneous, non-linear and lossy. We approximate both cases with (\ref{eq_wave_equation}), because data inaccuracies are inevitable and good for testing algorithmic robustness.

@@

Consider an acoustic signal transmitted to highlight a surface of interest. Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers. After stripping the carrier frequency by heterodyning and the signal signature by matched filtering (Fig. \ref{2_fig_article_relation}), an image is formed by coherently combining the receiver outputs to focus at one angle (or point) at a time. The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

@@

Imagine an array $\mathcal{A}$ with $M$ sensors at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length). Assume the array channels are digitally delayed to focus at $\udot{S}$, such that the delayed data from the $m$th channel at instant $n$ can be expressed as $x_m[n-\Delta_m, \vec{p}_{mS}]\in\mathbb{C}$. Focus at source $\udot{S}$ requires delays akin to
%
\begin{align}
\Delta_m[n] = \frac{\lVert \vec{p}_{mS}\rVert - \underset{m'}{\mathrm{min}}\ \lVert \vec{p}_{m'S}\rVert}{c},\label{2_eq_delta}
\end{align}
%
where both $\vec{p}_{mS}$ and $\vec{p}_{m'S}$ denote position vector $\udot{m}$ to $\udot{S}$ and subtracting the minimum distance is optional. By making the dependence on $\vec{p}$ implicit and assuming the data to be pre-delayed to the pixel of interest, we simplify the notation to $x_m[n]$. By definition, the beamformer output $z[n]\in\mathbb{C}$ is the weighted sum of delayed data samples:
%
\begin{align}
z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]} \in\mathbb{C},\label{2_eq_z}
\end{align}
%
where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. Spatial responses of the weights when the array is steered to broadside can be characterized by the beampattern,
%
\begin{align}
W(\vec{k}) = \sum_m^{M-1} w_m^* e^{j\vec k \cdot \vec{p}_m}.\label{2_eq_beampattern}
\end{align}
%
%
%
%
%. Beampattern is broadside-steered beamformer output in the presence of a single plane wave, as a function of the wave's angle.
%%Steered response is beamformer's output in the presence of some wave field, as a function of steering angle. 
%
The beamformer is called delay-and-sum (DAS) or conventional if the weights are static. A variety of weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}. A class of adaptive beamformers target this limitation by adapting the weights to the impinging wavefield. One such technique is the minimum variance distortionless response (MVDR) or Capon beamformer, named after its maker for use in spectral estimation~\cite{Capon1969}.

@@

\subsection{Time-domain MVDR}

@@

Time-domain MVDR often improves image contrast and resolution compared to conventional beamformers, as demonstrated in e.g. radar~\cite{Benitz1997}, ultrasound~\cite{Synnevag2007} and active sonar~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}. It achieves this by computing the weight set that minimizes noise and interference accumulated by the array, under the constraint of unit gain for signals from the pixel of interest. Assuming pre-delayed data, we can formulate this as
%
\begin{gather}
\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{a} = 1, \label{2_eq_mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
\end{gather}
%
where $\dvec{a}$ is steering vector. Applying Lagrange multipliers to this convex optimization problem yields the solution
%
\begin{gather}
\dvec w[n] = \frac{\Ri[n]\dvec{a}}{\dvec{a}^T\Ri[n]\dvec{a}} \qquad\text{where}\qquad \dvec{a}=\dvec{1}, \label{2_eq_weights}
\end{gather}
%
where $\dvec{1}$ is a row-vector of ones representing broadside phase-steering and $\R=E\{\dvec{x}[n]\dvec{x}^H[n]\}$ is the spatial covariance matrix for the full array. (\ref{2_eq_weights}) is the maximum likelihood estimate of the signal in the presence of zero-mean random Gaussian noise. We estimate $\R$ with a sample covariance matrix $\eR$, requiring some degree of
%
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
\end{itemize}%

@@

\subsection{Frequency-domain MVDR}

>>>>

% Related work can be found e.g. in ultrasound imaging, where Chen et. al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a}. Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix. 
% \\
\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
\caption{Method relations. This thesis explores three beamformers reconstructing images from acoustic echoes: element-space MVDR (reference), beam-space MVDR (close approximation) and element-space LCA (educated trial-and-error). Article IV presents background on synthesis of image templates from reconstructed images. }%
\label{2_fig_article_relation}
\end{figure}

@@

Any wave, regardless of its nature, is governed by a partial differential equation known as the \emph{wave equation}. Its form for a linear, homogeneous (translation-invariant) and isotropic (rotation-invariant) medium is
%
\begin{align}\label{eq_wave_equation}
\nabla^2 f 
 = \frac{1}{c}\frac{\delta^2 f}{\delta t^2},
\end{align}
%\frac{\delta^2 f}{\delta x^2} + \frac{\delta^2 f}{\delta y^2} + \frac{\delta^2 f}{\delta z^2}
%
where $\nabla^2$ is the Laplace operator, $f$ is field strength and $c$ is wave propagation speed. In complex mediums $c$ may depend on signal strength (non-linear), position (non-homogeneous/refractive), angle (anisotropic), frequency (dispersive) or be non-real (lossy). Deep sea is acoustically non-homogeneous due to varying pressure, temperature and salinity, but has negligible dispersion and attenuation. Biological tissue both non-homogeneous, non-linear and lossy. We approximate both cases with (\ref{eq_wave_equation}), because data inaccuracies are inevitable and good for testing algorithmic robustness.

@@

Consider an acoustic signal transmitted to highlight a surface of interest. Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers. After stripping the carrier frequency by heterodyning and the signal signature by matched filtering (Fig. \ref{2_fig_article_relation}), an image is formed by coherently combining the receiver outputs to focus at an angle or point at a time. The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

@@

Imagine an array $\mathcal{A}$ with $M$ sensors at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length). Assume the array channels are digitally delayed to focus at $\udot{S}$, such that the delayed data from the $m$th channel at instant $n$ can be expressed as $x_m[n-\Delta_m, \vec{p}_{mS}]\in\mathbb{C}$. Focus at source $\udot{S}$ requires delays akin to
%
\begin{align}
\Delta_m[n] = \frac{\lVert \vec{p}_{mS}\rVert - \underset{m'}{\mathrm{min}}\ \lVert \vec{p}_{m'S}\rVert}{c},\label{2_eq_delta}
\end{align}
%
where both $\vec{p}_{mS}$ and $\vec{p}_{m'S}$ denote position vector $\udot{m}$ to $\udot{S}$. Subtracting the minimum distance reduces latency and is optional. By making the dependence on $\vec{p}$ implicit and assuming the data to be pre-delayed to the pixel of interest, we simplify the notation to $x_m[n]$. By definition, the beamformer output $z[n]\in\mathbb{C}$ is a weighted sum of delayed data samples:
%
\begin{align}
z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]} \in\mathbb{C},\label{2_eq_z}
\end{align}
%
where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. Spatial responses of the weights for an array steered to broadside are characterized by the beampattern,
%
\begin{align}
W(\vec{k}) = \sum_m^{M-1} w_m^* e^{j\vec k \cdot \vec{p}_m}.\label{2_eq_beampattern}
\end{align}
%
%
%
%
%. Beampattern is broadside-steered beamformer output in the presence of a single plane wave, as a function of the wave's angle.
%%Steered response is beamformer's output in the presence of some wave field, as a function of steering angle. 
%
The beamformer is called delay-and-sum (DAS) or conventional if the weights are static. A variety of weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}. A class of adaptive beamformers target this limitation by adapting the weights to the impinging wavefield. One such technique is the minimum variance distortionless response (MVDR) or Capon beamformer, named after its maker for use in spectral estimation~\cite{Capon1969}.

@@

\subsection{Element-space MVDR}

@@

Element-space MVDR often improves image contrast and resolution compared to conventional beamformers~\cite{Benitz1997,Synnevag2007,Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}, but is computationally complex and statistically fragile. It seeks the weight set that minimizes noise and interference accumulated by the array, while enforcing unit gain for signals from the pixel of interest,
%
\begin{gather}
\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{a} = 1, \label{2_eq_mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
\end{gather}
%
where $\dvec{a}$ is steering vector. Applying Lagrange multipliers to this convex optimization problem yields the solution
%
\begin{gather}
\dvec w[n] = \frac{\Ri[n]\dvec{a}}{\dvec{a}^T\Ri[n]\dvec{a}} \qquad\text{where}\qquad \dvec{a}=\dvec{1}, \label{2_eq_weights}
\end{gather}
%
where $\dvec{1}$ is a row-vector of ones representing broadside phase-steering and $\R=E\{\dvec{x}[n]\dvec{x}^H[n]\}$ is the spatial covariance matrix for the full array. (\ref{2_eq_weights}) is the maximum likelihood estimate of the signal in the presence of zero-mean random Gaussian noise. We estimate $\R$ with a sample covariance matrix $\eR$, requiring some degree of
%
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Widrow1982,Kailath1985};
\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
\end{itemize}%

@@

\subsection{Beam-space MVDR}

### 2019-10-22 ThesisJo/3_publications.tex
2019-10-22

This article accelerates time-domain MVDR with GPUs for use in active sonar imaging. Time-domain MVDR is the reference method for creating sonar images using minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity. We refer to it as fully adaptive because it computes array weights analytically from spatial features in the data, yielding array weights from a continuous solution space that are optimal in the minimum variance sense. Like any data-driven algorithm, it requires statistically distinguishable signal and noise. However, in typical active systems, the signal is highly correlated with the noise and has poor temporal sample support. Therefore, in its unconstrained form, MVDR overfits the data, yields high variance and fails to generate representative images.

>>>>

This article accelerates element-space MVDR with GPUs for use in active sonar imaging. Element-space MVDR is the reference method for creating sonar images using minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity. We refer to it as fully adaptive because it computes array weights analytically from spatial features in the data, yielding array weights from a continuous solution space that are optimal in the minimum variance sense. Like any data-driven algorithm, it requires statistically distinguishable signal and noise. However, in typical active systems, the signal is highly correlated with the noise and has poor temporal sample support. Therefore, in its unconstrained form, MVDR overfits the data, yields high variance and fails to generate representative images.

### 2019-10-22 ThesisJo/4_discussion.tex
2019-10-22

\textbf{Paper I} - \textit{Time-domain MVDR beamforming on GPUs in active sonar imaging}:
\begin{enumerate}
\item Accelerates the MVDR beamformer through algorithmic reductions and fine-tuning to a GPU, yielding a real-time solution.
\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with no more than 32 channels, two orders of magnitude faster than a semi-optimized C-implementation on a CPU.
\item Offers rarely documented insights into hardware limitations, design implications and computational efficiency.
\end{enumerate}

\textbf{Paper II} - \textit{Frequency-domain MVDR beamforming on GPUs in clinical ultrasound imaging}:
\begin{enumerate}
\item Alters the implementation in Paper I to operate in beamspace (frequency domain), reduced to its three center beams (the most significant ones).
\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with 64 and 96 channels.
\item Yields similar image quality to time-domain MVDR, in accordance with former studies in active sonar.
\end{enumerate}

@@

GPU acceleration and algorithmic reductions largely mitigates the computational argument against MVDR in time-domain up to 32 channels and in time-domain for larger systems granted it is narrow-beam, narrow-band, or both. LCA is entirely data-parallel thus ideal for GPUs, but runs faster unoptimized on CPUs than accelerated MVDR.%  Real-time performance is likely possible.

>>>>

\textbf{Paper I} - \textit{Element-space MVDR beamforming on GPUs in active sonar imaging}:
\begin{enumerate}
\item Accelerates the MVDR beamformer through algorithmic reductions and fine-tuning to a GPU, yielding a real-time solution.
\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with no more than 32 channels, two orders of magnitude faster than a semi-optimized C-implementation on a CPU.
\item Offers rarely documented insights into hardware limitations, design implications and computational efficiency.
\end{enumerate}

\textbf{Paper II} - \textit{Beam-space MVDR beamforming on GPUs in clinical ultrasound imaging}:
\begin{enumerate}
\item Alters the implementation in Paper I to operate in beam-space (frequency domain), reduced to its three center beams (the most significant ones).
\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with 64 and 96 channels.
\item Yields similar image quality to element-space MVDR, in accordance with former studies in active sonar.
\end{enumerate}

@@

GPU acceleration and algorithmic reductions largely solves the issue with MVDR's computational complexity in element-space up to 32 channels, and in beam-space for larger systems that are narrow-beam, narrow-band, or both. LCA is entirely data-parallel thus ideal for GPUs, but runs faster unoptimized on CPUs than accelerated MVDR.%  Real-time performance is likely possible.

### 2019-10-22 ThesisJo/abstract.tex
2019-10-22

This thesis explores methods to accelerate adaptive acoustic image reconstruction and synthesis through beamforming and simulation, respectively. Beamforming is the main attraction: the art of delaying and weighting sensor arrays to craft desired spatial responses. When aiming to minimize image noise and interference, two good candidates are the minimum variance distortionless response (MVDR) and low complexity adaptive (LCA) beamformer. We study time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated guessing), and conclude that:
%
\begin{itemize}
\item \emph{Image quality} is similar for our adaptive beamformers, but improved upon compared to static methods.
%
% \item \emph{Adaptivity} is best for time-domain MVDR,
%
\item \emph{Acceleration} of MVDR through algorithmic reductions and implementation on graphics processing units (GPUs) yields real-time responses in time-domain up to 32 channels and in frequency-domain for any system size with narrow-band, narrow-beam, or both. LCA is inherently fast and extremely well suited for acceleration, 
\end{itemize}
%
Finally, we perform image synthesis of common objects on the seafloor with a GPU accelerated simulator. These are in turn used to perform adaptive automatic target recognition (ATR). The result is a more dynamic and flexible system yielding improved classification performance. Finally, we describe this system with a new navigational notation system, extended with topological concepts from robotics and homogeneous coordinates. It helps keep track of complex geometries, numerical representations and higher level abstractions such as manipulating rigid objects.

>>>>

This thesis explores methods to accelerate adaptive acoustic image reconstruction and synthesis through beamforming and simulation, respectively. Beamforming is the main attraction: the art of delaying and weighting sensor arrays to craft desired spatial responses. When aiming to minimize image noise and interference, two good candidates are the minimum variance distortionless response (MVDR) and low complexity adaptive (LCA) beamformer. We study element-space MVDR (reference), beam-space MVDR (close approximation) and element-space LCA (educated guessing), and conclude that:
%
\begin{itemize}
\item \emph{Image quality} is similar for our adaptive beamformers, but improved upon compared to static methods.
%
% \item \emph{Adaptivity} is best for time-domain MVDR,
%
\item \emph{Acceleration} of MVDR through algorithmic reductions and implementation on graphics processing units (GPUs) yields real-time responses in element-space up to 32 channels and in beam-space for any system size with narrow-band, narrow-beam, or both. LCA is inherently fast and extremely well suited for acceleration, 
\end{itemize}
%
Finally, we perform image synthesis of common objects on the seafloor with a GPU accelerated simulator. These are in turn used to perform adaptive automatic target recognition (ATR). The result is a more dynamic and flexible system yielding improved classification performance. Finally, we describe this system with a new navigational notation system, extended with topological concepts from robotics and homogeneous coordinates. It helps keep track of complex geometries, numerical representations and higher level abstractions such as manipulating rigid objects.
